---
title: "Fine-mapping"
author: "Siming Zhao"
date: '2024-01-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before the class

Install R package "susieR"

# Explore the data we will use the analysis
```{r}
rm(list=ls())
library(susieR)
data(N3finemapping)
attach(N3finemapping)
```

```{r}
names(N3finemapping)
```
The genotype matrix is `X`, The phenotype matrix is `Y`.  

We focus on the first trait, let

```{r}
y = Y[,1]
b = true_coef[,1]
which(b != 0)
```

Let's perform a univariate analysis

```{r}
sumstats <- univariate_regression(X, y)
z_scores <- sumstats$betahat / sumstats$sebetahat
log10p <- -log10(pchisq(z_scores^2,1,lower.tail=F))
susie_plot(z_scores,y="z",b=b)
```

# Fine mapping using individual level data

```{r}
fitted <- susie(X, y, L = 10)
```

By default, susie function computes 95% CS each containing one effect variable,
```{r}
print(fitted$sets)
```

Plot Posterior Inclusion Probability
```{r}
susie_plot(fitted, y="PIP", b=b, add_legend=T)
```

Choice of prior effect size: 

```{r}
fitted2 = susie(X, y, L = 10, estimate_prior_variance = FALSE, scaled_prior_variance = 0.2)
susie_plot(fitted2, y='PIP', b=b, add_legend=T)
```

```{r}
fitted2 = susie(X, y, L = 10, estimate_prior_variance = FALSE, scaled_prior_variance = 0.001)
susie_plot(fitted2, y='PIP', b=b, add_legend=T)
```

# Fine-mapping with summary statistics via susie_rss

z-scores are provided and we can compute R from X.

```{r}
R <- cor(X)
fitted_rss <- susie_rss(z_scores, R, L = 10)
plot(fitted$pip, fitted_rss$pip, ylim=c(0,1))
```

# Credit 

Credit to Gao Wang: https://statgenetics.github.io/statgen-courses/notebooks/finemapping.html#Fine-mapping-with-summary-statistics-via-susie_rss-6
