---
title: "Integrating eQTL and GWAS: TWAS and coloc"
author: "Siming Zhao"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before the class

Install R packages:

```{r, eval=F}
install.packages("remotes")
remotes::install_github("MRCIEU/TwoSampleMR")
```

Note: R >= 4.4 is required (MASS, a dependency, requires R >= 4.4 as of 2026). If you don't have R 4.4 or later, you can use [Posit Cloud](https://posit.cloud/).

**Important:** As of 2025, TwoSampleMR uses the [IEU OpenGWAS API](https://opengwas.io/) for data access, and API authentication is now required. Functions like `extract_instruments()`, `extract_outcome_data()`, `available_outcomes()`, and `clump_data()` will not work without a valid token. No local plink installation is needed — LD clumping is performed server-side.

To authenticate, register at [https://opengwas.io/](https://opengwas.io/) and set your token:

```{r, eval=FALSE}
token <- "YOUR_TOKEN_HERE"
ieugwasr::set_opengwas_jwt(token)
Sys.setenv(OPENGWAS_JWT = token)  # may be needed on some systems
```

Code blocks below that require API access are marked `eval=FALSE`. Run them interactively after setting your token.

# Perform MR

We will follow the [TwoSampleMR tutorial](https://mrcieu.github.io/TwoSampleMR/index.html) to perform a basic MR analysis. 

In this example, we will perform MR of BMI against coronary heart disease. To do so we need to identify the SNPs that influence the BMI (get instruments), and then extract those SNPs from a GWAS on coronary heart disease (effects of instruments on outcome), then perform MR.

```{r}
library(TwoSampleMR)
```

## Get instruments for exposures

First read in data for exposure: 

```{r}
bmi2_file <- system.file("extdata/bmi.csv", package = "TwoSampleMR")
bmi_exp_dat <- read_exposure_data(
    filename = bmi2_file,
    sep = ",",
    snp_col = "rsid",
    beta_col = "effect",
    se_col = "SE",
    effect_allele_col = "a1",
    other_allele_col = "a2",
    eaf_col = "a1_freq",
    pval_col = "p-value",
    units_col = "Units",
    gene_col = "Gene",
    samplesize_col = "n"
)
#> No phenotype name specified, defaulting to 'exposure'.
head(bmi_exp_dat)
```

To serve as valid instruments, they need to be independent. Thus we need to do clumping (those SNPs that have LD R-square above the specified threshold only the SNP with the lowest P-value will be retained). 

```{r, eval=FALSE}
# Requires API token — see authentication note above
bmi_exp_dat <- clump_data(bmi_exp_dat)
```

This package can also obtain data from thousands of existing GWAS directly.
A list of available GWAS:

```{r, eval=FALSE}
# Requires API token
ao <- available_outcomes()
head(ao)
```

To extract instruments for a particular trait using a particular study, for example to obtain SNPs for body mass index using the Locke et al 2015 GIANT study, you specify the study ID as follows:

This returns a set of LD clumped SNPs that are GWAS significant for BMI:
```{r, eval=FALSE}
# Requires API token
bmi2014_exp_dat <- extract_instruments(outcomes = 'ieu-a-2')
```

## Get effects of instruments on outcomes

We now need to find a suitable GWAS for coronary heart disease. We can search the available studies:

```{r, eval=FALSE}
# Requires API token
ao[grepl("heart disease", ao$trait), ]
```

The most recent CARDIOGRAM GWAS is ID number ieu-a-7. We can extract the BMI SNPs from this GWAS as follows:

```{r, eval=FALSE}
# Requires API token
chd_out_dat <- extract_outcome_data(
    snps = bmi_exp_dat$SNP,
    outcomes = 'ieu-a-7'
)
```

## Harmonise the exposure and outcome data

```{r, eval=FALSE}
dat <- harmonise_data(
    exposure_dat = bmi_exp_dat,
    outcome_dat = chd_out_dat
)
#> Harmonising Body mass index || id:ieu-a-2 (ieu-a-2) and Coronary heart disease || id:ieu-a-7 (ieu-a-7)
```

## Perform MR

We use two methods, MR-Egger and inverse variance weighting.
```{r, eval=FALSE}
res <- mr(dat, method_list = c("mr_egger_regression", "mr_ivw"))
res
```

# Analyze MR results

* Test for heterogeneity:

```{r, eval=FALSE}
mr_heterogeneity(dat)
```

* Horizontal pleiotropy

```{r, eval=FALSE}
mr_pleiotropy_test(dat)
```

* Scatter Plot

```{r, eval=FALSE}
p1 <- mr_scatter_plot(res, dat)
p1
```

